---
title: "Linear Regression in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
library(car)
```

# Linear Regression in R

Linear regression is used to model relationships between variables and make predictions.

## Simple Linear Regression

Modeling the relationship between two variables.

```{r simple-regression}
# Load data
data(mtcars)

# Fit simple linear regression: mpg ~ weight
model <- lm(mpg ~ wt, data = mtcars)
summary(model)
```

### Interpreting Results

- **Coefficients**: Intercept and slope
- **p-values**: Test if coefficients differ from 0
- **R²**: Proportion of variance explained
- **F-statistic**: Overall model significance

### Visualization

```{r simple-regression-plot}
# Create scatter plot with regression line
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Simple Linear Regression: MPG vs Weight",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon")
```

### Getting Coefficients

```{r coefficients}
# Extract coefficients
coef(model)

# Confidence intervals
confint(model)

# Predicted values
predict(model)

# Predict for new values
new_data <- data.frame(wt = c(2.5, 3.0, 3.5))
predict(model, new_data)
predict(model, new_data, interval = "confidence")  # CI for mean
predict(model, new_data, interval = "prediction")  # CI for prediction
```

## Multiple Linear Regression

Modeling relationships with multiple predictors.

```{r multiple-regression}
# Multiple regression: mpg ~ weight + hp + qsec
model_multiple <- lm(mpg ~ wt + hp + qsec, data = mtcars)
summary(model_multiple)
```

### Stepwise Selection

```{r stepwise}
# Full model
full_model <- lm(mpg ~ ., data = mtcars)

# Stepwise regression
step_model <- step(full_model, direction = "both")
summary(step_model)
```

## Model Diagnostics

Essential checks before interpreting results.

### 1. Linearity

```{r linearity}
# Residuals vs fitted
plot(model, which = 1)

# Component plus residual plot
crPlots(model)
```

### 2. Independence

```{r independence}
# Durbin-Watson test
library(lmtest)
dwtest(model)
```

### 3. Homoscedasticity (Equal Variance)

```{r homoscedasticity}
# Scale-location plot
plot(model, which = 3)

# Breusch-Pagan test
library(lmtest)
bptest(model)
```

### 4. Normality of Residuals

```{r normality}
# Q-Q plot of residuals
plot(model, which = 2)

# Histogram of residuals
hist(residuals(model), main = "Distribution of Residuals")

# Shapiro-Wilk test
shapiro.test(residuals(model))
```

### 5. Leverage and Influence

```{r leverage}
# Cook's distance
plot(model, which = 4)

# Leverage points
plot(model, which = 5)

# Identify influential points
influence.measures(model)

# Cook's distance values
cooksd <- cooks.distance(model)
which(cooksd > 4 / length(cooksd))  # Potential outliers
```

## All Diagnostic Plots

```{r all-diagnostics}
par(mfrow = c(2, 2))
plot(model)
```

## Correcting Violations

### Transformations

```{r transformations}
# Log transformation
model_log <- lm(mpg ~ log(wt), data = mtcars)
summary(model_log)

# Polynomial regression
model_poly <- lm(mpg ~ wt + I(wt^2), data = mtcars)
summary(model_poly)

# Box-Cox transformation
library(MASS)
bc <- boxcox(mpg ~ wt, data = mtcars)
```

### Handling Outliers

```{r outliers}
# Robust regression
library(MASS)
model_robust <- rlm(mpg ~ wt + hp, data = mtcars)
summary(model_robust)

# Compare with OLS
summary(lm(mpg ~ wt + hp, data = mtcars))
```

## Interaction Effects

```{r interactions}
# Model with interaction
model_int <- lm(mpg ~ wt * hp, data = mtcars)
summary(model_int)

# Visualize interaction
ggplot(mtcars, aes(x = wt, y = mpg, color = as.factor(cyl))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Interaction Effect")
```

## Dummy Variables

```{r dummy}
# Automatic encoding of factors
model_dummy <- lm(mpg ~ factor(cyl) + wt, data = mtcars)
summary(model_dummy)

# With reference level
mtcars$cyl_ref <- relevel(factor(mtcars$cyl), ref = "4")
model_ref <- lm(mpg ~ cyl_ref + wt, data = mtcars)
summary(model_ref)
```

## Model Comparison

```{r model-comparison}
# Compare models
model1 <- lm(mpg ~ wt, data = mtcars)
model2 <- lm(mpg ~ wt + hp, data = mtcars)
model3 <- lm(mpg ~ wt * hp, data = mtcars)

# AIC (lower is better)
AIC(model1, model2, model3)

# BIC
BIC(model1, model2, model3)

# ANOVA comparison
anova(model1, model2, model3)
```

## Cross-Validation

```{r cross-validation}
# Leave-one-out cross-validation
library(boot)

# CV function
cv.glm(mtcars, model_multiple)$delta

# k-fold CV
cv_10fold <- function(model, data, k = 10) {
  folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
  errors <- numeric(k)
  
  for (i in 1:k) {
    test_indices <- which(folds == i)
    train_data <- data[-test_indices, ]
    test_data <- data[test_indices, ]
    
    train_model <- lm(formula(model), data = train_data)
    predictions <- predict(train_model, test_data)
    errors[i] <- mean((test_data$mpg - predictions)^2)
  }
  
  return(mean(errors))
}

cv_10fold(model_multiple, mtcars)
```

## Predictive Accuracy

```{r accuracy}
# R-squared
summary(model_multiple)$r.squared

# Adjusted R-squared
summary(model_multiple)$adj.r.squared

# RMSE
sqrt(mean(residuals(model_multiple)^2))

# MAE
mean(abs(residuals(model_multiple)))
```

## Complete Example

```{r complete-example}
# Research question: What predicts fuel efficiency?

# Step 1: Explore relationships
pairs(mtcars[, c("mpg", "wt", "hp", "cyl", "disp")])

# Step 2: Fit initial model
model_full <- lm(mpg ~ ., data = mtcars)
summary(model_full)

# Step 3: Check assumptions
par(mfrow = c(2, 2))
plot(model_full)

# Step 4: Transform if needed
model_log <- lm(mpg ~ wt + hp, data = mtcars)
summary(model_log)

# Step 5: Final model
final_model <- lm(mpg ~ wt + hp + factor(cyl), data = mtcars)
summary(final_model)

# Step 6: Model diagnostics
shapiro.test(residuals(final_model))
bptest(final_model)
dwtest(final_model)

# Step 7: Report and interpret
cat("\n=== FINAL MODEL SUMMARY ===\n")
print(summary(final_model))
cat("\nR-squared =", round(summary(final_model)$r.squared, 3))
cat("\nAdjusted R-squared =", 
    round(summary(final_model)$adj.r.squared, 3))
```

## Summary

In this lesson, you learned:
- Simple and multiple linear regression
- Model diagnostics (assumptions)
- Handling violations
- Interactions and dummy variables
- Model comparison
- Cross-validation
- Predictive accuracy
- Complete regression workflow

## Next Steps

Learn advanced ANOVA techniques in the next lesson!

