---
title: "Introduction to R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to R

Welcome to R programming! This lesson will help you get started with R.

## What is R?

R is a programming language and environment for statistical computing and graphics. It's:
- **Free and open-source**
- **Powerful** for data analysis and visualization
- **Widely used** in academia and industry
- **Extensible** with thousands of packages

## Why Learn R?

- Data manipulation and analysis
- Statistical modeling
- Data visualization
- Reproducible research
- Open science

## The RStudio Interface

When you open RStudio, you'll see four panes:

1. **Source** (top left): Where you write and edit code
2. **Console** (bottom left): Where R executes commands
3. **Environment** (top right): Shows your objects and data
4. **Files/Plots/Help** (bottom right): Outputs and documentation

## Basic Operations

R works like a calculator:

```{r basic-calc}
2 + 3
10 - 4
5 * 6
20 / 4
```

## Variables and Assignment

Store values using the assignment operator `<-` or `=`:

```{r variables}
x <- 10
y <- 5
z <- x + y
print(z)
```

## Getting Help

R has extensive help documentation:

```{r help, eval=FALSE}
# Get help on any function
?print
help(print)

# Search for functions
??regression
```

## Data Types

R has several data types:

- **Numeric**: `5`, `3.14`
- **Character**: `"Hello"`, `"R"`
- **Logical**: `TRUE`, `FALSE`
- **Factor**: Categorical data
- And more...

## Useful Functions

```{r functions}
# Create a sequence
1:10

# Create a vector
c(1, 2, 3, 4, 5)

# Get the length
length(1:10)

# Get information about an object
class(1:10)
str(1:10)
```

## Working with Data

R can work with various data structures. Below are short, runnable examples that demonstrate how to create and use each of the main types: vectors, matrices, lists, and data frames.

### Vectors

Vectors are 1-dimensional atomic vectors (numeric, character, logical, etc.).

```{r vectors}
# numeric vector
v <- c(1, 2, 3, 4, 5)
v

# basic operations
v * 2
v[1]        # first element
length(v)   # length

# named vector
nv <- c(a = 10, b = 20, c = 30)
nv["b"]    # access by name
```

### Matrices

Matrices are 2-dimensional and hold elements of the same atomic type.

```{r matrices}
# create a 3x3 matrix
m <- matrix(1:9, nrow = 3, ncol = 3)
m

# index: row 1, all cols; all rows, col 2
m[1, ]
m[, 2]

# bind rows/columns
rbind(m, c(10, 11, 12))
cbind(m, c(10, 20, 30))
```

### Lists

Lists can contain heterogeneous elements and other objects (vectors, data.frames, functions).

```{r lists}
lst <- list(name = "Alice", age = 30, scores = c(90, 85, 88))
lst

# access by $ or [[ ]]
lst$name
lst[["age"]]
lst[["scores"]][2]  # second score
```

### Data frames

Data frames store tabular data where each column can be a different type. They're the primary data structure for most data analysis.

```{r dataframes}
df <- data.frame(
	id = 1:3,
	name = c("A", "B", "C"),
	score = c(88, 92, 75)
)
df

# column access
df$name

# subset rows (score > 80)
df[df$score > 80, ]

# add a new column
df$passed <- df$score >= 80
str(df)
summary(df)
```

These examples are small and intended to be run interactively in the console or knitted as part of this document. Play with the code: change values, try different indexing, and combine structures (for example, a list that contains a data frame).

## Practice Exercise

Try these in your R console:

1. Calculate your age in days
2. Create a variable with your name
3. Use the `seq()` function to create numbers from 1 to 20

## Next Steps

In the next lesson, we'll learn about different data types and structures in R.

## Summary

In this lesson, you learned:
- What R is and why it's useful
- The RStudio interface
- Basic mathematical operations
- How to create variables
- How to get help
- Basic data types and functions

