---
title: "Data Types and Structures in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Types and Structures

R has several data types and structures to organize and work with data.

## Vectors

A vector is a collection of elements of the same type.

### Numeric Vectors

```{r numeric-vector}
numbers <- c(1, 2, 3, 4, 5)
numbers

# Or use colon operator
1:5

# Or create a sequence
seq(1, 10, by = 2)
```

### Character Vectors

```{r character-vector}
names <- c("Alice", "Bob", "Charlie")
names

# Check the type
class(names)
```

### Logical Vectors

```{r logical-vector}
logical_vec <- c(TRUE, FALSE, TRUE, FALSE)
logical_vec
```

## Indexing and Subsetting

Access elements using square brackets:

```{r indexing}
vec <- c(10, 20, 30, 40, 50)

# Get first element
vec[1]

# Get multiple elements
vec[1:3]

# Get specific elements
vec[c(1, 3, 5)]

# Get all except some
vec[-1]
```

## Lists

Lists can contain different types of data:

```{r lists}
my_list <- list(
  name = "John",
  age = 30,
  scores = c(85, 90, 78)
)

my_list

# Access elements
my_list[[1]]
my_list$name
```

## Matrices

Matrices are two-dimensional arrays:

```{r matrices}
# Create a matrix
mat <- matrix(1:9, nrow = 3, ncol = 3)
mat

# Access elements
mat[1, 2]
mat[1, ]
mat[, 2]
```

## Data Frames

Data frames are like tables with rows and columns:

```{r dataframes}
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  score = c(85, 90, 78)
)

df

# Access columns
df$name
df[1, ]
df[, 2]
```

## Factors

Factors are used for categorical data:

```{r factors}
grades <- factor(c("A", "B", "A", "C", "B"))
grades

# Check levels
levels(grades)
```

## Checking Data Types

```{r check-types}
# Check if numeric
is.numeric(vec)

# Check if character
is.character(names)

# Check if data frame
is.data.frame(df)
```

## Practice Exercise

Create the following:
1. A numeric vector with numbers 1 to 10
2. A character vector with your favorite colors
3. A data frame with columns: name, age, city
4. Extract the age column from your data frame

## Summary

In this lesson, you learned:
- Vectors (numeric, character, logical)
- Lists
- Matrices
- Data frames
- Factors
- How to index and subset data
- How to check data types

